Staged Abstract Interpreters

### 2019-04-06 paper/10_intro.tex
paper

We show that these well-known abstractions can be combined smoothly, and more
importantly almost free. We implement the four interpreters using Scala and the
Lightweight Modular Staging \cite{DBLP:conf/gpce/RompfO10} framework. Through
staging, the major overhead of abstractions -- from the monadic layers -- is
completely eliminated in the generated code. We observe at most 14x times faster
than the unstage abstract interpreter. However, we do not lose any advantages of
this high-level programming style (e.g., modularity and extensibility); the
meaning of the program also remains the same, particularly, the soundness of
static analysis after staging.

@@

Furthermore, we show various applications of staging an abstract interpreter.
(1) We compare our approach with abstract compilation, and show that we can
achieve the same optimization by utilizing type-based stage annotations,
meanwhile, the analyzer program does not need any changes. (2) \todo{} (3)
Staging abstract interpreter enables modular analysis and provides a mechanized
way to create library summaries. We notice that we may apply the staged abstract
interpreter separately on the different part of the analyzed program, the
generated code is reusable and composable. Therefore, we can turn a
whole-program analyzer into an analyzer that runs modularly.

>>>>

We show that these well-known abstractions can be combined smoothly, and more
importantly almost free. We implement the four interpreters using Scala and the
Lightweight Modular Staging \cite{DBLP:conf/gpce/RompfO10} framework. Through
staging, the major overhead of abstractions -- from the monadic layers -- is
completely eliminated in the generated code. We observe up to 17x times faster
than the unstage abstract interpreter. However, we do not lose any advantages of
this high-level programming style (e.g., modularity and extensibility); the
meaning of the program also remains the same, particularly, the soundness of
static analysis after staging.

@@

Furthermore, we show various applications of staging an abstract interpreter.
(1) We compare our approach with abstract compilation, and show that we can
achieve the same optimization by utilizing type-based stage annotations,
meanwhile, the analyzer program does not need any changes. (2) We extend the basic staged abstract interpreter to diverse control-flow analysis, including store-widening analysis, context-sensitive analysis and abstract garbage collection.
(3) Staging abstract interpreter enables modular analysis and provides a mechanized
way to create library summaries. We notice that we may apply the staged abstract
interpreter separately on the different part of the analyzed program, the
generated code is reusable and composable. Therefore, we can turn a
whole-program analyzer into an analyzer that runs modularly.

### 2019-04-06 paper/20_prelim.tex
paper

\paragraph{Primitive Operations}
Other primitive operations can be implemented straightforwardly, for example,
@alloc@ simply takes the size of the current store and produce the successor number.
We elide most of them but describe a little bit on how do we handle functions
and applications as they will be very different when staging and abstract
interpretation is involved. The method @close@ takes a lambda term and an
environment, then create a closure value @CloV@.

>>>>

\paragraph{Primitive Operations}
Other primitive operations can be implemented straightforwardly, for example,
@alloc@ simply takes the size of the current store and produces the successor number.
We elide most of them but describe a little bit on how do we handle functions
and applications as they will be very different when staging and abstract
interpretation is involved. The method @close@ takes a lambda term and an
environment, then create a closure value @CloV@.

### 2019-04-06 paper/30_stageconc.tex
paper

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to obtain a staged concrete
interpreters by changing the monads and refactoring several primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala.

@@

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library that enables
dynamic code generation in a type-safe manner. Different from the approach of
MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} that
uses syntactic quotation and escape, LMS distinguishes binding-time solely based
on types. LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary
type, indicating a value of @T@ will be known at the next stage. All operations
acting a @Rep[T]@ expression will be residualized as generated code.

@@

For example, let us go back and see how LMS can be used to specialize the
@power@ function from Section ~\ref{intro}.

\todo{Show some IR}

@@

We use @RepStateT@ as an example to give a sketch of how the @Rep@ versions of monad
transformers look like. The generic type @M[_]@ is a @RepMonad@ and used as the
inner monad, which means the pair @(A,S)@ is also staged when appears inside of the monad
@M[_]@. The @flatMap@ is similar to the unstaged one, but the function @f@ takes
a @Rep[A]@ value. Again, @f@ is not a next-stage function, but a current-stage
function that transforms a next-stage value of @Rep[A]@ to @StateT[M,S,B]@.
\todo{highlight the changes in the code}

>>>>

In this section, based on the generic interpreter and concrete components we
presented in Section \ref{prelim}, we show how to obtain a staged concrete
interpreter by changing the monads and refactoring several primitive
operations. We begin by briefly introducing the Lightweight Modular Staging
framework in Scala.

@@

Lightweight modular staging (LMS) \cite{DBLP:conf/gpce/RompfO10} is a
multi-stage programming framework implemented as a Scala library that enables
dynamic code generation in a type-safe manner. Different from the approach of
MetaOCaml \cite{DBLP:conf/flops/Kiselyov14, DBLP:conf/gpce/CalcagnoTHL03} that
uses syntactic quotation and escape, LMS distinguishes binding-time solely based
on types. LMS provides a type constructor @Rep[T]@ where @T@ can be an arbitrary
type, indicating a value of @T@ will be known at the next stage. All operations
acting on @Rep[T]@ expressions will be residualized as generated code.

For the example we introduced in Section \ref{intro}, we have identified @b@ will 
be known later. To specialize it we write a function @snippet@ that provides a 
concrete value @x@ for @x@. After this, we may instantiate an @DslDriver@ object 
which provide the specialized program.

@@

@@

We use @RepStateT@ as an example to give a sketch of how the @Rep@ versions of monad
transformers look like. The generic type @M[_]@ is a @RepMonad@ and used as the
inner monad, which means the pair @(A,S)@ is also staged when appears inside of the monad
@M[_]@. The @flatMap@ is similar to the unstaged one, but the function @f@ takes
a @Rep[A]@ value. Again, @f@ is not a next-stage function, but a current-stage
function that transforms a next-stage value of @Rep[A]@ to @StateT[M,S,B]@.
%\todo{highlight the changes in the code}

### 2019-04-06 paper/abstract.tex
paper

As an example of abstraction without regret, we show that staging abstract
interpreters is practical and useful to optimize static analysis while requiring
least engineering efforts and not compromising any soundness. We conduct three
case studies, including a comparison with
\citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
the applicability on various control-flow analyses, and the use to modular
analysis. We also empirically evaluate the performance improved by staging. The
overhead of the abstraction layers is eliminated in the generated code, and the 
experiment shows an average speedup of 7x with staging for control-flow analysis.

>>>>

As an example of abstraction without regret, we show that staging abstract
interpreters is practical and useful to optimize static analysis while requiring
least engineering efforts and not compromising any soundness. We conduct three
case studies, including a comparison with
\citeauthor{Boucher:1996:ACN:647473.727587}'s abstract compilation,
the application of various control-flow analyses, and the use to modular
analysis. We also empirically evaluate the performance improved by staging. The
overhead of the abstraction layers is eliminated in the generated code, and the 
experiment shows an average speedup of 7x with staging for control-flow analysis.
